Microplatforms: Product platforms as code

### 2018-05-31 paper.tex

\begin{abstract}
	Microplatforms are an automation-based approach of provisioning and operating platforms to run digital products and services, which emphasises autonomy of cross-functional product teams. We discuss the traditional approach of building a large-scale, shared platform and explore the resulting cost to the organisation, chief of which is the cost of coordination between teams. Instead, we offer the concept of each cross-functional team owning and operating a separate, self-contained platform. A key enabler for this approach is full automation - capturing all aspects of the product and the platform from provisioning to service orchestration and policy as source code. This approach results in significantly increased team autonomy and enables agile architecture, continuous deployment to production and ongoing innovation. Microplatforms allow organisations to move engineering effort higher up the value chain and focus on using technology to solve customer problems.
\end{abstract}

@@

When building and delivering a real-world client-server application or service, it is important to consider all the parts that form the digital product, and without which the product would not operate. Typically we think of the product as two major components - the application itself and the \"platform\" it is hosted on. However, there are many other tools, processes, activities and concerns involved in delivering a Software as a Service (SaaS) product to its users than is covered by the functional features and the runtime environment alone (See figure \ref{fig:concerns}).

@@

Given the scope and complexity of all the activities we just described, it makes economic sense to design a set of processes and tools used to perform and support them and bundle them into a reusable, commodity solution we will refer to as a product platform. The effort involved should not be under-estimated~\cite{saaspdf24:online} and is the chief reason why most large organisations choose the approach of hosting their digital product portfolio on a single centrally managed shared platform. Most organisations also opt to run the product platform on cloud infrastructure, which has prooven to be more flexible and significantly cheaper to operate than the alternative of using on-premise infrastructure. (We will discuss the reasons for this cost reduction in more detail in sections~\ref{sec:cost} and~\ref{sec:automation}.)

There has been a clear trend over the past two decades of automating as much of the above activities as possible and treating them as another part of the product itself, managing them within the same teams as the product engineering. This is the core proposition of cloud services and a key concept of the DevOps movement~\cite{DevOpsCu34:online}. In this paper we will investigate ways of leveraging automation to design, build and manage everything involved in building and deploying digital product to the level where the entire product platform can be fully owned by a small, cross-functional product delivery team, enabling significantly increased capability and velocity of product delivery.

@@

In the long term, the platform should be flexible enough to support safe technical innovation in all directions and enable practices like evolutionary architecture~\cite{Microser21:online} and continuous improvement~\cite{KaizenTh68:online} even for the infrastructure. We will refer to these features using the term \textit{Agile Infrastructure}. Having the ability to evolve the infrastructure design eliminates the need for large up-front planning and enables ongoing innovation in a way that is similar to Agile software delivery.

@@

The initial investment involves the design, building, stabilising and scaling of the platform, typically done manually in the past, i.e. provisioning servers on premise or in the cloud, configuring them, linking them into HA clusters, setting up networking, etc. This requires specialist skills typically leading to the formation of a platform engineering team []. The cost of this initial phase always ends up significantly more than expected, understood or attributed.

@@

The overall cost of building and operating a shared product platform is large enough to pose a serious risk to the business. Typically, when things do not go well, this leads to resistance to cut the losses - the sunk cost fallacy~\cite{TheSunkC23:online}. Instead of trying again, or exploring different solutions, the attitude turns into \"we will make it work\" and results in products being built on a mediocre platform with reliability and scaling issues, increasing risk of preventable outages and slowing down product delivery. This is the final, often unrecognised, cost of a shared product platform - the missed opportunity cost of not delivering as much customer value as possible.

Even when things go well, the size of the investment means that when, eventually, the time comes to abandon the current platform in order to innovate using the latest available technology, nobody is willing to make the investment again, even though it could bring an order of magnitude improvement in efficiency of software delivery. This could be described as technology lock-in or a solution lock-in (similar to vendor lock-in []) - with a large-scale platform build, the required investment is so large that a single organisation is not likely to get a return on investment before the need to iterate on technology choices appears because the pace of technical innovation in the industry is so rapid.

It may seem like the only option left is to not spend the effort of building a shared platform in the first place and instead reach for a solution available in one of the cloud vendor offerings. This reduces the cost of building and scaling the platform, but the cost of sharing and the risk of lock-in stays. Further, the organisation is no longer fully in control of the available tools or the way they are being used by the teams and as a result, most organisations opt for some form of governance around the cloud offering as well, which effectively leads to the same situation as the custom build in terms of coordination cost, and the limited, centrally managed set of available features. Moving to a cloud native solution seems to treat some of the symptoms of the problem, but it does not cure the underlying cause, nor does it guarantee the end result will be any more fit for purpose.

@@

This is not a new finding, automating certain tasks is an accepted practice in multiple industries including software engineering. An obvious example is regression testing using automated unit tests, integration tests and end-to-end tests. Testing used to be a manual task performed by humans following test scripts. Automating it significantly decreased the time between deployments allowing us to get from deploying every few months to several times a day, enabling a completely new working practice known as Continuous Deployment to Production (CDP). We will focuse on the consequences of CDP at the end of section \ref{sec:microplatforms}.

@@

Second, \textbf{immutable infrastructure as code} enables automated, repeatable, reliable and fast infrastructure provisioning. All instances of infrastructure can be identical to the production environment. Since provisioning is automated, the infrastructure can be torn down and recreated within minutes. This lets us treat infrastructure as ephemeral, and allows us to update configuration by destroying and recreating, rather than modifying infrastructure in place, making it effectively immutable. This also means no access to servers is required, which simplifies configuration and increases security significantly.

@@

A container is a unit of deployment agnostic of the underlying infrastructure and the technology stack used by the application inside it. This draws a very strong boundary between the infrastructure and the application. The images are described with Dockerfiles, recipes of creating images step by step, stored as code []. Docker images are formed of layers, which are immutable and content addressed (by a cryptographic hash of their content) []. These properties mean that Docker images can be built once and deployed on any infrastructure supporting Docker.

@@

With most cluster orchestration tools (e.g. Docker in Swarm mode, Kubernetes) sets of configuration, we will refer to them as “stacks” (a name used by Docker), can be deployed under an alias, which also forms a namespace for the services. Within the namespace, services can refer to each other by local names, making the entire stack portable, without any change of configuration. This gives a really light-weight mechanism of creating logical environments in a cluster.

As an added benefit, the service orchestration is immutable. In order to change the configuration or topology of the microservices, the code describing it is changed and the orchestration redeployed. The cluster orchestration system then reconciles the differences so that only the parts that are different actually change. Repeated deployments and updates are idempotent - if there is no difference, nothing changes. This declarative approach means we can logically treat the whole service orchestration as a single unit for deployment and let the automation handle the details. We show the specifics of making sure this property holds in Appendix~\ref{sec:architecture}.

Service orchestration as code also enables a practice that has become known as GitOps, which advocates performing operations in a runtime environment through changes to a version control repository. In practice, it can mean a cluster has an associated git repository holding all the service orchestration configuration for every single application running in it. In order to deploy an application, a CI service pushes a new revision into this repository and the cluster (more precisely a service running within it) responds to the event by pulling the configuration and attempting to deploy it. This serves as an inversion of control mechanism, removing the need for a CI service (typically in a non-production environment) to have direct access to and control over clusters in a production environment, and also allows for final checks to be performed by the cluster itself to decide whether to accept the deployment. As a side-effect, the state of the cluster is fully audited over time and can be restored in a different cluster even months later. Optionally, this also means no outside access to deployment APIs is needed for the production clusters, increasing security.

In figure \ref{fig:concerns} we show 48 cross-cutting concerns, identified by Sean Treadway~\cite{Treadway17:online}, that need addressing for each microservice. None of them are directly related to adding customer value. With microservices architecture, we have made a rod for our own back: developers should be able to write code that directly adds customer value and they should not need to worry about any of the concerns listed. These can be addressed by employing a service mesh (such as Istio []). It allows us to declaratively specify \textbf{policy as code} providing services such as intelligent routing and load balancing, traffic shaping and shifting, security policies, ingress and egress rules, circuit breaking [] and upstream call retries, rate limits, call tracing, telemetry and reporting. It is even possible to use a service mesh to inject faults into the system to test reliability features that may have been overlooked. All of this configuration is declarative and stored as code alongside the orchestration configuration.

@@

For truly stateless microservices, it is simpler if each of the applications has its own copy of a microservice treating it simply as a module of the system, re-using and sharing only its source code. In order to share a microservice instance, it needs to be fully treated as a service, as it is no longer under the control of the team managing the application - it essentially becomes a 3rd party dependency. In that case, it should have an agreed service level, support model, documentation, versioning and migration strategy, etc.

@@

It removes the coordination cost of all product teams depending on a single platform team. If a team needs to change some aspect of their platform or add a missing feature, they can do so for themselves. A platform team will most likely still exist, but the product it delivers is the design and automation of a platform, not the infrastructure itself. This team can become an owner and steward of the platform design and automation code, which can and should be internally open source, so that everyone can contribute and the platform engineers can transition into expert support and innovation roles, rather than day to day operations. This is a tried and tested approach which works well for shared, reusable code libraries and services []. With the shared dependency removed, the organisation becomes more horizontally scalable, just like software systems do. Vertical integration leads to horizontal scalability.

@@

Targeting a cluster orchestration technology, applications become fully portable between any infrastructure supporting the technology from a developer’s laptop through any public cloud provider to on-premise solutions. In the past year, Kubernetes [] has rapidly become a de-facto standard abstraction in cluster orchestration: it is possible to run a local cluster using a tool like Minikube [], clusters can be provisioned using simple compute resources in different clouds with Kops [], all three major cloud providers are offering or have announced hosted services (GKE [], EKS [] and AKS []) and managed on-premise solutions also exist, e.g. OpenShift by RedHat []or Pivotal Kubernetes Service [].

@@

With continuous deployment to production, every change to the application and infrastructure goes through thorough automated testing and validation. Once the validation passes without issues, the change is immediately deployed to production. This can happen several times a day, even multiple times an hour.

@@

All of the above - automation, immutability, idempotence and autonomy are cases of simplicity, \textit{simple} being distinctly different from \textit{easy} [simple made easy]. By simple, we mean having a minimum of structure, little coupling and entanglement, avoiding N:M relationships between entities. Immutability is also a case of decoupling - mutable state is one of the hardest to understand couplings - a coupling through time.

@@

Repeat the 5 things that enable Microplatforms
Repeat the 3 process/engineering approaches that Microplatforms
Go through how this applies to the goals stated in the introduction

>>>>

\begin{abstract}
	Microplatforms are an automation-based approach to provisioning and operating platforms running digital products and services, which emphasises autonomy of cross-functional product teams. We discuss the traditional approach of building a large-scale, shared platform and explore the resulting cost to the organisation, chief of which is the cost of coordination between teams. Instead, we offer the concept of each cross-functional team owning and operating a separate, self-contained platform. A key enabler for this approach is full automation - capturing all aspects of the product and the platform from provisioning to service orchestration and policy as source code. This approach results in significantly increased team autonomy and enables agile architecture, continuous deployment to production and ongoing innovation. Microplatforms allow organisations to move engineering effort higher up the value chain and focus on using technology to solve customer problems.
\end{abstract}

@@

When building and delivering a real-world client-server application or service, it is important to consider all the parts that form the digital product, and without which the product would not operate. Typically we think of the product as two major components - the application itself and the ``platform'' it is hosted on. However, there are many other tools, processes, activities and concerns involved in delivering a Software as a Service (SaaS) product to its users than is covered by the functional features and the runtime environment alone (See figure \ref{fig:concerns}).

@@

Given the scope and complexity of all the activities we just described, it makes economic sense to design a set of processes and tools used to perform and support them and bundle them into a reusable, commodity solution we will refer to as a product platform. The effort involved should not be under-estimated~\cite{saaspdf24:online} and is the chief reason why most large organisations choose the approach of hosting their digital product portfolio on a single centrally managed shared platform. Most organisations also opt to run the product platform on cloud infrastructure, which has proven to be more flexible and significantly cheaper to operate than the alternative of using on-premise infrastructure. (We will discuss the reasons for this cost reduction in more detail in sections~\ref{sec:cost} and~\ref{sec:automation}.)

There has been a clear trend over the past two decades of automating as many of the above activities as possible and treating them as another part of the product itself, managing them within the same teams as the product engineering. This is the core proposition of cloud services and a key concept of the DevOps movement~\cite{DevOpsCu34:online}. In this paper we will investigate ways of leveraging automation to design, build and manage everything involved in building and deploying digital product to the level where the entire product platform can be fully owned by a small, cross-functional product delivery team, enabling significantly increased capability and velocity of product delivery.

@@

In the long term, the platform should be flexible enough to support safe technical innovation in all directions and enable practices like evolutionary architecture~\cite{Microser21:online} and continuous improvement~\cite{KaizenTh68:online} even for the infrastructure. We will refer to these features using the term \textit{Agile Infrastructure}. Having the ability to evolve the infrastructure design reduces the need for up-front planning and enables ongoing innovation in a way that is similar to Agile software delivery.

@@

The initial investment involves the design, building, stabilising and scaling of the platform, typically done manually in the past, i.e. provisioning servers on premise or in the cloud, configuring them, linking them into highly-available clusters, setting up networking, etc. This requires specialist skills typically leading to the formation of a platform engineering team []. The cost of this initial phase always ends up significantly more than expected, understood or attributed.

@@

The overall cost of building and operating a shared product platform is large enough to pose a serious risk to the business. Typically, when things do not go well, this leads to resistance to cut the losses - the sunk cost fallacy~\cite{TheSunkC23:online}. Instead of trying again, or exploring different solutions, the attitude turns into ``we will make it work'' and results in products being built on a mediocre platform with reliability and scaling issues, increasing risk of preventable outages and slowing down product delivery. This is the final, often unrecognised, cost of a shared product platform - the missed opportunity cost of not delivering as much customer value as possible.

Even when things go well, the size of the investment means that when, eventually, the time comes to abandon the current platform in order to innovate using the latest available technology, nobody is willing to make the investment again, even though it could bring an order of magnitude improvement in efficiency of software delivery. This could be described as technology lock-in or a solution lock-in (similar to vendor lock-in []) - with a large-scale platform build, the required investment is so large that a single organisation is not likely to get a return on investment before the need to iterate on technology choices appears because the pace of technical innovation in the industry is ever accelerating.

It may seem like the only option left is to not spend the effort of building a shared platform in the first place and instead reach for a solution available in one of the cloud vendor offerings. This reduces the cost of building and scaling the platform, but the cost of sharing and the risk of lock-in stays. Furthermore, the organisation is no longer fully in control of the available tools or the way they are being used by the teams and as a result, most organisations opt for some form of governance around the cloud offering as well, which effectively leads to the same situation as the custom build in terms of coordination cost, and the limited, centrally managed set of available features. Moving to a cloud native solution seems to treat some of the symptoms of the problem, but it does not cure the underlying cause, nor does it guarantee the end result will be any more fit for purpose.

@@

This is not a new finding, automating certain tasks is an accepted practice in multiple industries including software engineering. An obvious example is regression testing using automated unit tests, integration tests and end-to-end tests. Testing used to be a manual task performed by humans following test scripts. Automating it significantly decreased the time between deployments allowing us to get from deploying every few months to several times a day, enabling a completely new working practice known as Continuous Deployment to Production (CDP). We will focus on the consequences of CDP at the end of section \ref{sec:microplatforms}.

@@

Second, \textbf{immutable infrastructure as code} enables automated, repeatable, reliable and fast infrastructure provisioning. All instances of infrastructure can be identical to the production environment. Since provisioning is automated, the infrastructure can be torn down and recreated within minutes. This lets us treat infrastructure as ephemeral, and allows us to update configuration by destroying and recreating, rather than modifying infrastructure in place, making it effectively immutable. Because no in-place modification is allowed, no human access to servers is required, which simplifies configuration and increases security significantly.

@@

A container is a unit of deployment agnostic of the underlying infrastructure and the technology stack used by the application inside it. This draws a very strong boundary between the infrastructure and the application. The images are described with Dockerfiles, recipes for creating images step by step, stored as code []. Docker images are formed of layers, which are immutable and content addressed (by a cryptographic hash of their content) []. These properties mean that Docker images can be built once and deployed on any infrastructure supporting Docker.

@@

With most cluster orchestration tools (e.g. Docker in Swarm mode, Kubernetes) sets of configuration, we will refer to them as “stacks” (a name used by Docker), can be deployed under an alias, which also forms a namespace for the services. Within the namespace, services can refer to each other by local names, making the entire stack portable, without any change of configuration. This gives a really light-weight mechanism for creating logical environments in a cluster.

As an added benefit, the service orchestration is immutable. In order to change the configuration or topology of the microservices, the code describing it is changed and the orchestration redeployed as a logical unit (as if the application was monolithic). The cluster orchestration system then reconciles the differences so that only the parts that are different actually change. Repeated deployments and updates are idempotent - if there is no difference, nothing changes. This declarative approach means we can logically treat the whole service orchestration as a single unit for deployment and let the automation handle the details. We show the specifics of making sure this property holds in Appendix~\ref{sec:architecture}.

Service orchestration as code also enables a practice that has become known as GitOps [ref], which advocates performing operations in a runtime environment through changes to a version control repository. In practice, it can mean a cluster has an associated git repository holding all the service orchestration configuration for every single application running in it. In order to deploy an application, a CI service pushes a new revision into this repository and the cluster (more precisely a service running within it) responds to the event by pulling the configuration and attempting to deploy it. This serves as an inversion of control mechanism, removing the need for a CI service (typically in a non-production environment) to have direct access to and control over clusters in a production environment, and also allows for final checks to be performed by the cluster itself to decide whether to accept the deployment. As a side-effect, the state of the cluster is fully audited over time and can be restored in a different cluster even months later. Optionally, this also means no outside access to deployment APIs is needed for the production clusters, increasing security.

In figure \ref{fig:concerns} we show 48 cross-cutting concerns, identified by Sean Treadway~\cite{Treadway17:online}, that need addressing for each microservice. None of them are directly related to adding customer value. With microservices architecture, we have made a rod for our own back: developers should be able to write code that directly adds customer value and they should not need to worry about any of the concerns listed. These can be addressed by employing a service mesh (such as Istio []). It allows us to declaratively specify \textbf{policy as code} providing services such as intelligent routing and load balancing, traffic shaping and shifting, security policies, ingress and egress rules, circuit breaking [] and upstream call retries, rate limits, call tracing, telemetry and reporting. It is even possible to use a service mesh to inject faults into the system to test reliability features that may have been overlooked (similar concept to Netflix's Chaos Monkey []). All of this configuration is declarative and stored as code alongside the orchestration configuration.

@@

For stateless microservices, it is simpler if each of the applications has its own copy of a microservice treating it simply as a module of the system, re-using and sharing only its source code, not the running instance(s). In order to share a microservice instance, it needs to be fully treated as a service, as it is no longer under the control of the team managing the application - it essentially becomes a 3rd party dependency. In that case, it should have an agreed service level, support model, documentation, versioning and migration strategy, etc.

@@

It removes the coordination cost of all product teams depending on a single platform team. If a team needs to change some aspect of their platform or add a missing feature, they can do so for themselves. This is a fundamental shift in the scope of developers' responsibility for their application - they are not only responsible for the product features but also for reliably delivering the service to their customers (i.e. true DevOps).

A platform team will most likely still exist, but the product it delivers is the design and automation of a platform, not the infrastructure itself. This team can become an owner and steward of the platform design and automation code, which can and should be internally open source, so that everyone can contribute and the platform engineers can transition into expert support and innovation roles, rather than day to day operations. This is a tried and tested approach which works well for shared, reusable code libraries and services []. With the shared dependency removed, the organisation becomes more horizontally scalable, just like software systems do. Vertical integration leads to horizontal scalability.

@@

Targeting a cluster orchestration technology, applications become fully portable between any infrastructure supporting the technology from a developer’s laptop through any public cloud provider to on-premise solutions. In the past year, Kubernetes [] has rapidly become a de-facto standard abstraction in cluster orchestration: it is possible to run a local cluster using a tool like Minikube [], clusters can be provisioned using simple compute resources in different clouds with Kops [], all three major cloud providers are offering hosted services (GKE [], EKS [] and AKS []) and managed on-premise solutions also exist, e.g. OpenShift by RedHat []or Pivotal Kubernetes Service [].

@@

With continuous deployment to production, every change to the application \textit{and} infrastructure goes through thorough automated testing and validation. Once the validation passes without issues, the change is immediately deployed to production. This can happen several times a day, even multiple times an hour.

@@

All of the above - automation, immutability, idempotence and autonomy are cases of simplicity, \textit{simple} being distinctly different from \textit{easy} [simple made easy]. By simple, we mean having a minimum of structure, little coupling and entanglement, avoiding N:M relationships between entities. Immutability is also a case of decoupling - mutable state is one of the hardest couplings to understand - a coupling through time.

@@

The overall goal of a product platform should be to minimise engineering effort spent on anything but delivering value to customers. This covers a vast array of capabilities and as a consequence, designing and building a product platform requires a large up-front and ongoing investment. The traditional way of offseting this cost and ensuring consistency is to build a single, shared, centrally managed product platform to host the entire product portfolio and serve all product delivery teams. We discussed how this approach leads to a growth in scope of the platform features and large coordination costs. The overall effort required can very easily grow to a point where the platform project fails entirely to deliver, or does not bring any return on investment in time, before it is necessary to iterate on the underlying technology choices, requiring yet more investment.

As an alternative, we suggest a different approach focused on enabling autonomy of product delivery teams. It is based on high levels of automation, managing the platform design in its entirety as an internal open source software project, which can be contributed to by everyone, adapted for individual requirements and can evolve with the needs of the organisation and its customers.

The central concept of the approach is to manage the application, build, testing and deployment pipelines, service orchestration, runtime policies and infrastructure provisionig as code. The full automation leads to repeatability, consistency of environments and enables safe Continuous Deployment to Production. Adopting solutions based on principles of simplicity, immutability and idempotence reduces the complexity of understanding and effort required to the point of allowing full ownership of their platform by every product delivery team. With the shared platform team dependency removed, the engineering organisation becomes more horizontally scalable and agile and a larger portion of the engineering effort can be focused on delivering value to customers.
